Visual Diagnostics for Constrained Optimisation with Application to Guided Tours

### 2021-03-23 paper/zhang-cook-laa-langrene-menendez.tex
proper citation of ferrn

\abstract{%
Guided tour searches for interesting low-dimensional views of
high-dimensional data via optimising a projection pursuit index
function. The first paper of projection pursuit by
\citet{friedman1974projection} stated that ``the technique used for
maximising the projection index strongly influences both the statistical
and the computational aspects of the procedure.'' While much work has
been done in proposing indices in the literature, less has been done on
evaluating the performance of the optimisers. In this paper, we
implement a data collection object in the optimisation of projection
pursuit guided tour and introduce visual diagnostics based on the data
object collected. These diagnostics and this workflow can be applied to
a broad class of optimisers, to assess their performance. An R package,
\pkg{ferrn}, has been created to implement the diagnostics.

@@

Visualisation is widely used in exploratory data analysis
\citep{tukey1977exploratory, unwin2015graphical, healy2018data, wilke2019fundamentals}.
Presenting information in graphics often unveils information that would
otherwise not be discovered and provides a more comprehensive
understanding of the problem at hand. Task specific tools such as
\citet{li2020visualizing} show how visualisation can be used to
understand, for instance, the behaviour of neural network on
classification models. However, no general visualisation tool is
available for diagnosing optimisation procedures. The work presented in
this paper brings visualization tools into optimisation problems with
the aim to better understand the performance of optimisers in practice.

The goal of continuous optimisation is to find the best solution within
the space of all feasible solutions where typically the best solution is
decided by an objective function. Broadly speaking, optimization can be
unconstrained or constrained \citep{kelley1999iterative}. The
unconstrained problem can be formulated as a minimization (or
maximization) problem such as \(\min_{x} f(x)\) where
\(f:\mathbb{R}^n \rightarrow \mathbb{R}\) is an objective function with
certain properties defined in an \(L^p\) space. In this case, solutions
rely on gradient descent or ascent methods. In the constrained
optimization problem additional restrictions are introduced via a set of
functions that can be convex or non-convex:
\(g_i:\mathbb{R}^n \rightarrow \mathbb{R}\) for \(i = 1, \ldots k\) and
hence the problem can be written as \(\min_{x} f(x)\) \emph{subject to}
\(g_i(x) \leq 0\). Here methods such as multipliers and convex
optimization methods including linear and quadratic programming can be
used.

The focus of this paper is on the optimisation problem arising in the
projection pursuit guided tour \citep{buja2005computational}, an
exploratory data analysis tool used for detecting interesting structures
in high-dimensional data through a set of lower-dimensional projections
\citep{cook2008grand}. The goal of the optimisation is to identify the
projection, characterised by the projection basis, that gives the most
interesting low-dimensional view. A view is interesting if it can show
some structures of the data, e.g., bi-modality, clustering and outliers,
other than normal distribution.

The optimization challenges encountered in the projection pursuit guided
tour problem are common to those of optimization in general. Examples of
those include the existence of multiple optima (local and global), the
trade off between computational burden and proximity to the optima,
dealing with noisy objective functions that might be non-smooth and
non-differentiable \citep{jones1998efficient}. Those are not unique to
this context and therefore the visualization tools and optimization
methods presented in this paper can be easily applied to any other
optimization problems.

The remainder of the paper is organised as follows. Section \ref{optim}
provides an overview of optimisation methods, specifically line search
methods. Section \ref{tour} reviews projection pursuit guided tour,
defines the optimisation problem and outlines three existing algorithms.
Section \ref{vis-diag} presents the new visual diagnostics. A data
structure is defined to capture information during the optimisation, and
different diagnostic plots are designed with the data collected. Section
\ref{application} shows applications of how these diagnostic plots can
be used to discover interesting aspects of different optimisers and
guide improvement to the existing algorithms. Finally, Section
\ref{implementation} describes the R package: \pkg{ferrn}, that
implements the visual diagnostics.

@@

A large class of methods utilises the gradient information of the
objective function, with the most notable one being the gradient ascent
(descent) method. Although gradient optimization methods are popular,
they rely on the availability of the objective function derivatives and
on the complexity of the constraints. As will be shown in the next
section, the independent variable, in our optimisation problem, is a
matrix of projection basis and the computational time required to
perform differentation on a matrix could impede the render of tour
animation. Hence, gradient-based methods are not the focus of this paper
and considerations will be given to derivative-free methods.

Derivative-free methods, which do not rely on the knowledge of the
gradient, are more generally applicable. Derivative-free methods have
been developed over the years, where the emphasis is on finding, in most
cases, a near optimal solution. Below will introduce three
derivative-free methods, two of which are random search methods:
creeping random search and simulated annealing, and the other one is
pseudo derivative search.

Random search mehtods \citep{zabinsky2013stochastic} contain a random
sampling component in its algorithm and has shown its ability to
optimise non-convex and non-smooth functions. The initial random search
algorithm, pure random search \citep{Brooks1958-zk}, draws candidate
points from the entire space without using any information of the
current position and updates the current position when an improvement on
the objective function is made. As the dimension of the space becomes
larger, sampling from the entire space would indicate a long time for
convergence to occur despite a guaranteed global convergence
\citep{spall2005introduction}. Various algorithms have then been
developed to improve pure random search by either concentrating on a
narrower sampling space or using a different updating machanism.
Creeping random search \citep{White1971-ci} is such a variation where a
candidate point is generated within a neighbourhood of the current
point. This gives creeping random search an advantage of being fast to
compute, however, global convergence is not longer guaranteed. On the
other hand, simulated annealing
\citep{kirkpatrick1983optimization, bertsimas1993simulated}, takes a
different updating machanism from the pure random search. Rather than
only updating the current point when an improvement is made, simulated
annealing uses a Metropolis acceptance criterion, where nonimproving
candidates can still have a chance to be accepted. The convergence of
simulated annealing algorithms has been widely researched
\citep{mitra1986convergence, belisle1992convergence, granville1994simulated}
and the global optimum can be attained under mild conditions.

Pseudo derivative search uses a common search scheme in optimistaion:
line search. In line search methods, users are required to provide an
initial estimate \(x_{1}\) and, at each iteration, a search direction
\(S_k\) and a step size \(\alpha_k\) are generated. Then one moves on to
the next point following \(x_{k+1} = x_k + \alpha_kS_k\) and the process
is repeated until the desire convergence is reached. In derivative-free
methods, local information of the objective function are used to
determine the search direction. The choice of step size also needs
considerations, as inadequate step sizes might prevent the optimisation
method to converge to an optimum. An ideal step size can be chosen via
finding the value of \(\alpha_k \in \mathbb{R}\) that maximises
\(f(x_k + \alpha_kS_k)\) with respect to \(\alpha_k\) at each iteration.

@@

Let \(\mathbf{X}_{n \times p}\) be the data matrix with \(n\)
observations in \(p\) dimensions. A \(d\)-dimensional projection is a
linear transformation from \(\mathbb{R}^p\) into \(\mathbb{R}^d\)
defined as \(\mathbf{Y} = \mathbf{X} \cdot \mathbf{A}\), where
\(\mathbf{Y}_{n \times d}\) is the projected data and
\(\mathbf{A}_{p\times d}\) is the projection basis matrix. We define
\(f: \mathbb{R}^{n \times d} \mapsto \mathbb{R}\) to be an index
function that maps the projected data \(\mathbf{Y}\) onto a scalar
value. This is commonly known as the projection pursuit index function,
or just index function, and is used to measure the ``interestingness''
of a given projection. An interesting projection shows structures that
are non-normal and this has been argued for two reasons: Theoretical
proofs from \citet{diaconis1984asymptotics} show that projections tend
to be normal, as \(n\) and \(p\) approaching infinity under certain
conditions. Index functions proposed in the litearture include early
indices that can be categorised by measuring the \(L^2\) distance
between the projection and a normal distribution: Legendre index
\citep{friedman1974projection}, Hermite index
\citep{hall1989polynomial}, and natural Hermite index
\citep{cook1993projection}; chi-square index \citep{posse1995projection}
for detecting spiral structure; kurtosis index \citep{Loperfido2020} and
skewness index \citep{Loperfido2018} for detecting outliers in financial
time series; and most recently, scagnostic indices \citep{laa2020using}
for summarising structures in scatterplot matrices based on eight
scagnostic measures.

As a general visualisation method, tour produces animations of high
dimensional data via rotations of low-dimensional planes. There are
different tour implementations depending on how the high dimensional
space is investigated: grand tour \citep{cook2008grand} selects the
planes randomly to provide a general overview; manual tour
\citep{cook1997manual} gradually phases in and out one variable to
understand the contribution of that variable in the projection. Guided
tour, the main interest of this paper, chooses the planes with the aid
of projection pursuit, to gradually reveal the most interesting
projection. Given a random start, projection pursuit iteratively finds
bases with higher index values and the guided tour constructs a geodesic
interpolation between these planes to form a tour path. Figure
\ref{fig:tour-path} shows a sketch of the tour path where the blue
frames are produced by the projection pursuit optimisation and the white
frames are interpolations between the blue frames. Mathematical details
of the geodesic interpolation can be found in
\citet{buja2005computational}. The aforementioned tour method has been
implemented in the R package \CRANpkg{tourr} \citep{tourr}.

@@

In projection pursuit the optimisation problem aims at finding the
global and local maxima that give interesting projections according to
an index function. That is, it starts with a given randomly selected
basis \(\mathbf{A}_1\) and aims at finding an optimal final projection
basis \(\mathbf{A}_T\) that satisfies the following optimisation
problem:

@@

\noindent where \(f\) and \(\mathbf{X}\) are defined as in the previous
section, \(\mathcal{A}\) is the set of all \(p\)-dimensional projection
bases, \(I_d\) is the \(d\)-dimensional identity matrix, and the
constraint ensures the projection bases, \(\mathbf{A}\), to be
orthonormal. It is worth noticing the following: 1) The optimisation is
constrained and the orthonormality constraint imposes a geometrical
structure on the bases space: it forms a Stiefel manifold. 2) There may
be index functions for which the objective function might not be
differentiable. 3) While finding the global optimum is the goal of the
optimisation problem, interesting projections may also appear in the
local optimum. 4) The optimisation should be computationally agile since
the tour animation is viewed by the users during the optimisation.

@@

Three optimisers have been implemented in the \CRANpkg{tourr}
\citep{tourr} package: Creeping random search (CRS), simulated annealing
(SA), and pseudo derivative (PD). Creeping random search (CRS) is a
random search optimiser that samples a candidate basis
\(\mathbf{A}_{l}\) in the neighbourhood of the current basis
\(\mathbf{A}_{\text{cur}}\) by
\(\mathbf{A}_{l} = (1- \alpha)\mathbf{A}_{\text{cur}} + \alpha \mathbf{A}_{\text{rand}}\)
where \(\alpha\) controls the radius of the sampling neighbourhood and
\(\mathbf{A}_{\text{rand}}\) is generated randomly. \(\mathbf{A}_{l}\)
is then orthonormalised to fulfil the basis constraint. If
\(\mathbf{A}_{l}\) has an index value higher than the current basis
\(\mathbf{A}_{\text{cur}}\), the optimiser outputs \(\mathbf{A}_{l}\)
for guided tour to construct an interpolation path. The neighbourhood
parameter \(\alpha\) is adjusted by a cooling parameter:
\(\alpha_{j+1} = \alpha_j * \text{cooling}\) before the next iteration
starts. The optimiser terminates when the maximum number of iteration
\(l_{\max}\) is reached before a better basis can be found. The
algorithm of SA is summarised in Algorithm \ref{random-search} in the
appendix. \citet{posse1995projection} has proposed a slightly different
cooling scheme by introducing a halving parameter \(c\). In his
proposal, the \(\alpha\) is only adjusted if the last iteration takes
more than \(c\) times to find a better basis.

Simulated annealing (SA)
\citep{kirkpatrick1983optimization, bertsimas1993simulated} uses the
same sampling process as CRS but allows a probabilistic acceptance of a
basis with lower index value than the current. Given an initial value of
\(T_0\), the temperature at iteration \(l\) is defined as
\(T(l) = \frac{T_0}{\log(l + 1)}\). When a candidate basis fails to have
an index value larger than the current basis, SAJO gives it a second
chance to be accepted with probability
\[P= \min\left\{\exp\left[-\frac{\mid I_{\text{cur}} - I_{l} \mid}{T(l)}\right],1\right\}\]
where \(I_{(\cdot)}\) denotes the index value of a given basis. This
implementation allows the optimiser to make a move and explore the basis
space even the landing basis does have a higher index value and hence
enable the optimiser to jump out of a local optimum. The algorithm
\ref{simulated-annealing} in the appendix highlights how SAJO differs
from SA in the inner loop.

Pseudo derivative (PD) search \citep{cook1995grand} uses a different
strategy than SA and SAJO. Rather than randomly sample the basis space,
PD first computes a search direction by evaluating vases close to the
current basis. A step size is then chosen along the geodesic direction
by another optimisation over an 90 degree angle from \(-\pi/4\) to
\(\pi/4\). The resulting candidate basis \(\mathbf{A}_{**}\) is returned
for the current iteration if it has a higher index value than the
current one. Algorithm \ref{search-geodesic} in the appendix summarises
the inner loop of the PD.

@@

A starting point of diagnosing an optimiser is to understand how many
searches an optimiser has conducted. One may want to simply plot the
index value of the search points across its natural order, but a point
geometry may work well if each iteration has a similar number of points.
When some iterations have considerably more points than others, using a
point geometry over-emphasizes the iterations that have more search
points since these iterations will occupy the mast majority of the plot
space. An alternative is to summarise the search in each iteration using
boxplots and each iteration will then be spaced out equally.
Occasionally, one may still want to switch back to a point geometry if
the number of points is small in a particular iteration and this can be
achieved via the \texttt{cutoff} argument in the search plot function.
Additional annotations are added to facilitate better reading of the
plot and these include\\
1) the number of points searched in each iteration can be added as text
label at the bottom of each iteration; 2) the anchor bases to
interpolate are connected and highlighted in a larger size; and 3) the
colour of the last iteration is in a grey scale to indicate no better
basis found in this iteration.

Figure \ref{fig:toy-search} shows an example of the search plot for CRS
(left) and SA (right). Both optimisers quickly find better bases in the
first few iterations and then take longer to find one in the later
iterations. The anchor bases, the ones found with the highest index
value in each iteration, always have an increased index value in the
optimiser CRS while this is not the case for SA. This feature gives SA
an advantage in this simple example to quickly find the optimum.

@@

Apart from checking the search and progression of an optimiser, looking
at where the bases are positioned in the basis space is another
interest. Given the orthonormality constraint, the space of projection
bases \(\mathbf{A}_{p \times d}\) is a \(p \times d\)-dimensional
sphere. A dimensionality reduction method, e.g.~principal component
analysis is applied to first project all the bases onto a 2D space. In a
projection pursuit guided tour optimisation, there are various types of
bases involved: 1) The starting basis; 2) The search bases that the
optimiser evaluated to produce the anchor bases; 3) The anchor bases
that have the highest index value in each iteration; 4) The
interpolating bases on the interpolation path; and finally 5) the end
basis. The importance of these bases differs and the most important ones
are the starting, interpolating and end bases. The anchor and search
bases can be turned on with argument \texttt{details\ =\ TRUE}.
Sometimes, two optimisers can start with the same basis but finish with
bases of opposite sign. This happens because the projection is invariant
to the sign difference of the bases and so does the index value,
however, this creates difficulties for comparing the optimisers since
the end bases will be symmetric with respect to the origin. A sign
flipping step is conducted to flip the signs of all the bases in one
routine if different optimisations finish at opposite places.

@@

While the previous few examples have looked at the space of 1D bases in
a unit sphere, this section visualises the space of 2D bases. Recall
that the columns in a 2D basis are orthogonal to each other, the space
of \(p \times 2\) bases is a \(p\)-D torus. If take \(p = 3\), one would
see a classical 3D torus shape as shown by the grey points in Figure
\ref{fig:toy-torus}. The two circles of the torus can be observed to be
perpendicular to each other and this can be linked back to the
orthogonality of the columns in the bases. Two paths optimised by CRS
and PD are plotted on top of the torus and coloured in green and brown
respectively to match the previous few 1D space plots. The final basis
found by PD and CRS are shown in a larger shape and printed below
respectively:

@@

The implementation of this projection has been divided into two
packages: the data collection object is implemented in the existing CRAN
package \CRANpkg{tourr} \citep{tourr} while the optimiser diagnostics
have been implemented in a new package, \pkg{ferrn}. When a guided tour
is run, the users can choose to collect the data by binding
\texttt{animate\_*()} to a variable. Once the data object has been
obtained, the \pkg{ferrn} package provides diagnostic plots shown in
Section \ref{vis-diag}. The main functions of the \pkg{ferrn} package
functionality is listed below.

>>>>

\abstract{%
Guided tour searches for interesting low-dimensional views of
high-dimensional data via optimising a projection pursuit index
function. The first paper of projection pursuit by
\citet{friedman1974projection} stated that ``the technique used for
maximising the projection index strongly influences both the statistical
and the computational aspects of the procedure.'' While much work has
been done in proposing indices in the literature, less has been done on
evaluating the performance of the optimisers. In this paper, we
implement a data collection object in the optimisation of projection
pursuit guided tour and introduce visual diagnostics based on the data
object collected. These diagnostics and this workflow can be applied to
a broad class of optimisers, to assess their performance. An R package,
\CRANpkg{ferrn} \citep{ferrn}, has been created to implement the
diagnostics.

@@

Visualisation is widely used in exploratory data analysis
\citep{tukey1977exploratory, unwin2015graphical, healy2018data, wilke2019fundamentals}.
Presenting information in graphics often unveils insights that would
otherwise not be discovered and provides a more comprehensive
understanding of the problem at hand. Task specific tools such as
\citet{li2020visualizing} show how visualisation can be used to
understand, for instance, the behaviour of the optimisation for the
example of neural network classification models. However, no general
visualisation tool is available for diagnosing optimisation procedures.
The work presented in this paper brings visualization tools into
optimisation problems with the aim to better understand the performance
of optimisers in practice.

The goal of continuous optimisation is to find the best solution within
the space of all feasible solutions where typically the best solution is
defined via an objective function. Broadly speaking, optimization can be
unconstrained or constrained \citep{kelley1999iterative}. The
unconstrained problem can be formulated as a minimization (or
maximization) problem such as \(\min_{x} f(x)\) where
\(f:\mathbb{R}^n \rightarrow \mathbb{R}\) is an objective function with
certain properties defined in an \(L^p\) space. In this case, solutions
rely on gradient descent or ascent methods. In the constrained
optimization problem additional restrictions are introduced via a set of
functions that can be convex or non-convex:
\(g_i:\mathbb{R}^n \rightarrow \mathbb{R}\) for \(i = 1, \ldots k\) and
hence the problem can be written as \(\min_{x} f(x)\) \emph{subject to}
\(g_i(x) \leq 0\). Here methods such as multipliers and convex
optimization methods including linear and quadratic programming can be
used.

The focus of this paper is on the optimisation problem arising in the
projection pursuit guided tour \citep{buja2005computational}, an
exploratory data analysis tool used for detecting interesting structures
in high-dimensional data through a set of lower-dimensional projections
\citep{cook2008grand}. The goal of the optimisation is to identify the
projection, represented by the projection matrix, that gives the most
interesting low-dimensional view. A view is interesting if it can show
some structures of the data other than a normal distribution, e.g.,
bi-modality, clustering and outliers.

The optimization challenges encountered in the projection pursuit guided
tour problem are common to those of optimization in general. Examples
include the existence of multiple optima (local and global), the
trade-off between computational burden and proximity to the optima, or
dealing with noisy objective functions that might be non-smooth and
non-differentiable \citep{jones1998efficient}. The visualization tools
and optimization methods presented in this paper can therefore easily be
applied to other optimization problems.

The remainder of the paper is organised as follows. Section \ref{optim}
provides an overview of optimisation methods, specifically line search
methods. Section \ref{tour} reviews projection pursuit guided tour,
defines the optimisation problem and outlines three existing algorithms.
Section \ref{vis-diag} presents the new visual diagnostics. A data
structure is defined to capture information during the optimisation, and
different diagnostic plots are designed with the data collected. Section
\ref{application} shows applications of how these diagnostic plots can
be used to discover interesting aspects of different optimisers and
guide improvements to the existing algorithms. Finally, Section
\ref{implementation} describes the R package: \CRANpkg{ferrn}
\citep{ferrn}, that implements the visual diagnostics.

@@

A large class of methods uses the gradient information of the objective
function, with the most notable one being the gradient ascent (descent)
method. Although gradient optimization methods are popular, they rely on
the availability of the objective function derivatives and on the
limited complexity of the constraints. As will be shown in the next
section, the independent variables in our optimisation problem are the
entries of a projection matrix and the computational time required to
perform differentiation on a matrix could impede the render of tour
animation. Hence, gradient-based methods are not the focus of this paper
and considerations will be given to derivative-free methods.

Derivative-free methods, which do not rely on the knowledge of the
gradient, are more generally applicable. Derivative-free methods have
been developed over the years, where the emphasis is on finding, in most
cases, a near optimal solution. Here we consider three derivative-free
methods, two of which are random search methods: creeping random search
and simulated annealing, and the other one is pseudo derivative search.

Random search methods \citep{zabinsky2013stochastic} have a random
sampling component as part of their algorithm and have been shown to
have the ability to optimise non-convex and non-smooth functions. The
initial random search algorithm, pure random search
\citep{Brooks1958-zk}, draws candidate points from the entire space
without using any information of the current position and updates the
current position when an improvement on the objective function is made.
As the dimension of the space becomes larger, sufficient sampling from
the entire space would require a long time for convergence to occur,
despite a guaranteed global convergence \citep{spall2005introduction}.
Various algorithms have thus been developed to improve pure random
search by either concentrating on a narrower sampling space or using a
different updating machanism. Creeping random search
\citep{White1971-ci} is such a variation, where a candidate point is
generated within a neighbourhood of the current point. This makes
creeping random search faster to compute, however, global convergence is
not longer guaranteed. On the other hand, simulated annealing
\citep{kirkpatrick1983optimization, bertsimas1993simulated}, introduces
a different updating mechanism. Rather than only updating the current
point when an improvement is made, simulated annealing uses a Metropolis
acceptance criterion, where worse candidates still have a chance to be
accepted. The convergence of simulated annealing algorithms has been
widely researched
\citep{mitra1986convergence, belisle1992convergence, granville1994simulated}
and the global optimum can be attained under mild conditions.

Pseudo derivative search uses a common search scheme in optimistaion:
line search. In line search methods, users are required to provide an
initial estimate \(x_{1}\) and, at each iteration, a search direction
\(S_k\) and a step size \(\alpha_k\) are generated. Then one moves on to
the next point following \(x_{k+1} = x_k + \alpha_kS_k\) and the process
is repeated until the desired convergence is reached. In derivative-free
methods, local information of the objective function is used to
determine the search direction. The choice of step size also needs
consideration, as inadequate step sizes might prevent the optimisation
method to converge to an optimum. An ideal step size can be chosen by
finding the value of \(\alpha_k \in \mathbb{R}\) that maximises
\(f(x_k + \alpha_kS_k)\) with respect to \(\alpha_k\) at each iteration.

@@

Let \(\mathbf{X}_{n \times p}\) be the data matrix with \(n\)
observations in \(p\) dimensions. A \(d\)-dimensional projection is a
linear transformation from \(\mathbb{R}^p\) into \(\mathbb{R}^d\)
defined as \(\mathbf{Y} = \mathbf{X} \cdot \mathbf{A}\), where
\(\mathbf{Y}_{n \times d}\) is the projected data and
\(\mathbf{A}_{p\times d}\) is the projection matrix. We define
\(f: \mathbb{R}^{n \times d} \mapsto \mathbb{R}\) to be an index
function that maps the projected data \(\mathbf{Y}\) onto a scalar
value. This is commonly known as the projection pursuit index function,
or just index function, and is used to measure the ``interestingness''
of a given projection. An interesting projection shows structures that
are non-normal and this has been argued for two reasons: Theoretical
proofs from \citet{diaconis1984asymptotics} show that projections tend
to be normal, as \(n\) and \(p\) approach infinity under certain
conditions. Index functions proposed in the litearture include early
indices that can be categorised by measuring the \(L^2\) distance
between the projection and a normal distribution: Legendre index
\citep{friedman1974projection}, Hermite index
\citep{hall1989polynomial}, and natural Hermite index
\citep{cook1993projection}; chi-square index \citep{posse1995projection}
for detecting spiral structure; kurtosis index \citep{Loperfido2020} and
skewness index \citep{Loperfido2018} for detecting outliers in financial
time series; and most recently, scagnostic indices \citep{laa2020using}
for summarising structures in scatterplot matrices based on eight
scagnostic measures.

As a general visualisation method, tour produces animations of high
dimensional data via rotations of low-dimensional planes. There are
different versions depending on how the high dimensional space is
investigated: grand tour \citep{cook2008grand} selects the planes
randomly to provide a general overview; manual tour
\citep{cook1997manual} gradually phases in and out one variable to
understand the contribution of that variable in the projection. Guided
tour, the main interest of this paper, chooses the planes with the aid
of projection pursuit, to gradually reveal the most interesting
projection. Given a random start, projection pursuit iteratively finds
bases with higher index values and the guided tour constructs a geodesic
interpolation between these planes to form a tour path. Figure
\ref{fig:tour-path} shows a sketch of the tour path where the blue
frames are produced by the projection pursuit optimisation and the white
frames are interpolations between the blue frames. Mathematical details
of the geodesic interpolation can be found in
\citet{buja2005computational}. The aforementioned tour method has been
implemented in the R package \CRANpkg{tourr} \citep{tourr}.

@@

In projection pursuit the optimisation aims at finding the global and
local maxima that give interesting projections according to an index
function. That is, it starts with a given randomly selected basis
\(\mathbf{A}_1\) and aims at finding an optimal final projection basis
\(\mathbf{A}_T\) that satisfies the following optimisation problem:

@@

\noindent where \(f\) and \(\mathbf{X}\) are defined as in the previous
section, \(\mathcal{A}\) is the set of all \(p\)-dimensional projection
bases, \(I_d\) is the \(d\)-dimensional identity matrix, and the
constraint ensures the projection bases, \(\mathbf{A}\), to be
orthonormal. It is worth noticing the following: 1) The optimisation is
constrained and the orthonormality constraint imposes a geometrical
structure on the bases space: it forms a Stiefel manifold. 2) There may
be index functions for which the objective function might not be
differentiable. 3) While finding the global optimum is the goal of the
optimisation problem, interesting projections may also appear in the
local optimum. 4) The optimisation should be fast to compute since the
tour animation is viewed by the users during the optimisation.

@@

Three optimisers have been implemented in the \CRANpkg{tourr}
\citep{tourr} package: Creeping random search (CRS), simulated annealing
(SA), and pseudo derivative (PD). Creeping random search (CRS) is a
random search optimiser that samples a candidate basis
\(\mathbf{A}_{l}\) in the neighbourhood of the current basis
\(\mathbf{A}_{\text{cur}}\) by
\(\mathbf{A}_{l} = (1- \alpha)\mathbf{A}_{\text{cur}} + \alpha \mathbf{A}_{\text{rand}}\)
where \(\alpha\) controls the radius of the sampling neighbourhood and
\(\mathbf{A}_{\text{rand}}\) is generated randomly. \(\mathbf{A}_{l}\)
is then orthonormalised to fulfil the basis constraint. If
\(\mathbf{A}_{l}\) has an index value higher than the current basis
\(\mathbf{A}_{\text{cur}}\), the optimiser outputs \(\mathbf{A}_{l}\)
for guided tour to construct an interpolation path. The neighbourhood
parameter \(\alpha\) is adjusted by a cooling parameter:
\(\alpha_{j+1} = \alpha_j * \text{cooling}\) before the next iteration
starts. The optimiser terminates when the maximum number of iteration
\(l_{\max}\) is reached before a better basis can be found. The
algorithm of CRS is summarised in Algorithm \ref{random-search} in the
appendix. \citet{posse1995projection} has proposed a slightly different
cooling scheme by introducing a halving parameter \(c\). In his
proposal, the \(\alpha\) is only adjusted if the last iteration takes
more than \(c\) times to find a better basis.

Simulated annealing (SA)
\citep{kirkpatrick1983optimization, bertsimas1993simulated} uses the
same sampling process as CRS but allows a probabilistic acceptance of a
basis with lower index value than the current. Given an initial value of
\(T_0\), the temperature at iteration \(l\) is defined as
\(T(l) = \frac{T_0}{\log(l + 1)}\). When a candidate basis fails to have
an index value larger than the current basis, SA gives it a second
chance to be accepted with probability
\[P= \min\left\{\exp\left[-\frac{\mid I_{\text{cur}} - I_{l} \mid}{T(l)}\right],1\right\}\]
where \(I_{(\cdot)}\) denotes the index value of a given basis. This
implementation allows the optimiser to make a move and explore the basis
space even if the candidate basis does not have a higher index value and
hence enable the optimiser to jump out of a local optimum. The algorithm
\ref{simulated-annealing} in the appendix highlights how SA differs from
CRS in the inner loop.

Pseudo derivative (PD) search \citep{cook1995grand} uses a different
strategy than CRS and SA. Rather than randomly sample the basis space,
PD first computes a search direction by evaluating bases close to the
current basis. A step size is then chosen along the corresponding
geodesic by another optimisation over an 90 degree angle from \(-\pi/4\)
to \(\pi/4\). The resulting candidate basis \(\mathbf{A}_{**}\) is
returned for the current iteration if it has a higher index value than
the current one. Algorithm \ref{search-geodesic} in the appendix
summarises the inner loop of the PD.

@@

A starting point of diagnosing an optimiser is to understand how many
searches it has conducted, i.e.~we to summarise how the index is
increasing over iterations and how many tries need to be sampled at each
iteration. This is achieved using the search plot function: a boxplot
shows the distribution of index values for each try, where the selected
basis (corresponding to the highest index value) is always shown as a
point. In case of few tries at a given iteration showing the data points
directly may be preferred over the boxplot, this is controlled via the
\texttt{cutoff} argument. Additional annotations are added to facilitate
better reading of the plot and these include 1) the number of points
searched in each iteration can be added as text label at the bottom of
each iteration; 2) the anchor bases to interpolate are connected and
highlighted in a larger size; and 3) the colour of the last iteration is
in a grey scale to indicate no better basis found in this iteration.

Figure \ref{fig:toy-search} shows an example of the search plot for CRS
(left) and SA (right). Both optimisers quickly find better bases in the
first few iterations and then take longer to find one in the later
iterations. The anchor bases, the ones found with the highest index
value in each iteration, always have an increased index value in the
optimiser CRS while this is not the case for SA. This feature gives CRS
an advantage in this simple example to quickly find the optimum.

@@

Apart from checking the search and progression of an optimiser, looking
at where the bases are positioned in the basis space is also of
interest. Given the orthonormality constraint, the space of projection
bases \(\mathbf{A}_{p \times d}\) is a Stiefel manifold. For
one-dimensional projections this forms a \(p\)-dimensional sphere. A
dimensionality reduction method, e.g.~principal component analysis is
applied to first project all the bases onto a 2D space. In a projection
pursuit guided tour optimisation, there are various types of bases
involved: 1) The starting basis; 2) The search bases that the optimiser
evaluated to produce the anchor bases; 3) The anchor bases that have the
highest index value in each iteration; 4) The interpolating bases on the
interpolation path; and finally 5) the end basis. The importance of
these bases differs and the most important ones are the starting,
interpolating and end bases. The anchor and search bases can be turned
on with argument \texttt{details\ =\ TRUE}. Sometimes, two optimisers
can start with the same basis but finish with bases of opposite sign.
This happens because the projection is invariant to orientation of the
basis and so is the index value. However, this creates difficulties for
comparing the optimisers since the end bases will be symmetric with
respect to the origin. A sign flipping step is conducted to flip the
signs of all the bases in one routine if different optimisations finish
at opposite places.

@@

While the previous few examples have looked at the space of 1D bases in
a unit sphere, this section visualises the space of 2D bases. Recall
that the columns in a 2D basis are orthogonal to each other, so the
space of \(p \times 2\) bases is a \(p\)-D torus. For \(p = 3\) one
would see a classical 3D torus shape as shown by the grey points in
Figure \ref{fig:toy-torus}. The two circles of the torus can be observed
to be perpendicular to each other and this can be linked back to the
orthogonality. Two paths from CRS and PD are plotted on top of the torus
and coloured in green and brown respectively to match the previous
plots. The final basis found by PD and CRS are shown in a larger shape
and printed below respectively:

@@

This project contributes to the software development in two packages: a
data collection object is implemented in \CRANpkg{tourr} \citep{tourr},
while the visual diagnostics of the optimisers is implemented in
\CRANpkg{ferrn} \citep{ferrn}. Below list functions in the
\CRANpkg{ferrn} \citep{ferrn} package:
